---
title: "Statistical Learning Comparisons"
author: "Kodyak"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS) # for LDA lda()
library(class) # for NearestNeighbors knn()
library(knitr) # for kable
library(boot) # for bootstrap
```

### Problems


1.  Evaluate the methods of classification we have discussed using class-validation for the data at pages.uoregon.edu/~dlevin/DATA/ozo.txt.  The response variance is HO, which indicates whether a given day had high ozone levels or not.  Discuss the success or failures of these methods.
 

2. The standard errors in linear regression are based on the assumption that the errors are independent, have the same variance, and have a Normal distribution.

(a) Simulate data where the errors are not Normal, e.g. come from a t-distribution with 3 degrees of freedom.  (The t-distribution is like the normal, but has larger probabilities of extreme data points.)  (Use rt in R to simulate.  Assume the model is y = 3 + 3x + error, where there are 100 x's spread between 0 and 3.)

(b) Fit a linear model, and compare the standard errors of the fitted coefficients to the standard errors estimated by the bootstrap method.

(c) Now simulate where the errors are correlated. (We will discuss in class how to do that.)  Compare the standard errors reported by the regression output (for the coefficients) to the bootstrap-estimated standard errors.


.....

.....


1. We will examine the classification methods: Logistic Regression, LDA (Linear Discriminant Analysis), and $k-$Nearest Neighbors on the ozo dataset (classifying whether days had high or low ozone levels). We will compare the average error rate of cross-validated model fitting from each area, to compare relative success rates of the techniques. In particular, we optimize the error rate in $k-$Nearest Neighbors with respect to choice of neighbor size. For details, see the associated R markdown file.

```{r initialize_data, echo=FALSE, warning=FALSE}

#import and view data
ozo = read.table(file="ozo.txt",header=TRUE,row.names=1)
attach(ozo)

#plot(vh,wind,col=humidity,pch=19) #looking at data


set.seed(2) # for analysis reproducibility in case of random neighbor tie - breaking
# Note: possible that a bad seed will cause a confusion matrix to bcomee degenerate
# if all predictions are only one of TRUE or FALSE. 
# Due to how it was coded, this can pass a subtle error.


#The first step is to figure out how to divide into k parts. 
#There are many ways to do this. 
#One trick is to randomize the order of the rows, and then 
#take contiguous chunks:
n = dim(ozo)[1]
ozor = ozo[sample(1:n,n,replace=FALSE),]
k = 10 #cross validation order
j.breaks = seq(1,by = I(n/k),length.out=I(k+1))
j.breaks[I(k+1)] = n
train.d = vector("list",k); test.d = vector("list",k)
for(i in 1:k){
  jj = j.breaks[i]:j.breaks[i+1]
  test.d[[i]] = ozor[jj,]
  train.d[[i]] = ozor[-jj,] 
}

#  initialize error vector
errorRate_list <- 1:k 

```

```{r logistic_regression, echo=FALSE, warning=FALSE}

## Logistic Regression 

for(i in 1:k){
 
  train.frame <- train.d[[i]] #train on    train.d[i]
  test.frame <- test.d[[i]]
  
#  attach(train.frame)   #unnecessary
  
  #build model
  logmodel <- glm( HO ~ . , data = train.frame, family=binomial)
  #model_list[i] <- logmodel
  
  # test on   test.frame
  predictions <- round(predict(logmodel, newdata = test.frame, type="response"))
  
    for(j in 1:length(predictions)) {
      if (predictions[j] == 1) {predictions[j] = "TRUE"} else { predictions[j] = "FALSE"}
    } 
  
  confusionMatrix <- table(predictions, test.frame$HO)
  #print(confusionMatrix)
  
  errorRate_list[i] <- (confusionMatrix[2] + confusionMatrix[3])/(sum(confusionMatrix))
  # confusion matrix is a list of 2-dim vectors, this picks out the off diagonal (error) terms
  
}

#overall error rate of logistic regression 
errorRate_logistic <- sum(errorRate_list)/k

```

```{r lda, echo=FALSE, warning=FALSE}

## LDA 

for(i in 1:k){
  
  train.frame <- train.d[[i]] #train on    train.d[i]
  test.frame <- test.d[[i]]
  
  #  attach(train.frame)   #unnecessary
  
  #build model
  lda.fit <- lda( HO ~ . , data = train.frame)
  #model_list[i] <- logmodel
  
  # test on   test.frame
  lda.pred=predict(lda.fit, test.frame)

  predictions <- lda.pred$class # class = predictions from the lda model predict() function
  
  confusionMatrix <- table(predictions, test.frame$HO)
  #print(confusionMatrix)
  
  errorRate_list[i] <- (confusionMatrix[2] + confusionMatrix[3])/(sum(confusionMatrix))
  # confusion matrix is a list of 2-dim vectors, this picks out the off diagonal (error) terms
  
}

#overall error rate of LDA 
errorRate_lda <- sum(errorRate_list)/k

```

```{r knn, echo=FALSE, warning=FALSE}

## K Nearest Neighbors 

    #OPTIMISING KNN
    MaxNeighborSize = 25
    errorRate_N = 1:MaxNeighborSize # vector of KNN error rates to find optimal neighbor size
    
    for(N in 1:MaxNeighborSize){
      neighbor = N
      for(i in 1:k){
        
        train.frame <- train.d[[i]] #train on    train.d[i]
        test.frame <- test.d[[i]]
        
        predictions = knn(train.frame, test.frame, train.frame$HO , k=neighbor)
        
        confusionMatrix <- table(predictions, test.frame$HO)
        #    print(confusionMatrix)
        
        errorRate_list[i] <- (confusionMatrix[2] + confusionMatrix[3])/(sum(confusionMatrix))
        # confusion matrix is a list of 2-dim vectors, this picks out the off diagonal (error) terms
        
      }
      errorRate_N[N] <- sum(errorRate_list)/k
    }
    inputs = 1:MaxNeighborSize
    errors = errorRate_N
    
    error_v_neighbor_plot <- plot(x = inputs, y = errors, ylim=c(0.14,.2), main = "Average Error Rate vs Neighbor Size", xlab = "neighbor size" , ylab = "average error rate")  # visual inspection of error vs neighbor size.

    
    # optimalSize reports the minimal error choice.
    optimalSize <- inputs[which.min(errors)]

# Compute KNN predictions

# Fix number of nearest neighbors 'neighbor' 
neighbor = optimalSize

for(i in 1:k){
  
  train.frame <- train.d[[i]] #train on    train.d[i]
  test.frame <- test.d[[i]]
  
  predictions = knn(train.frame, test.frame, train.frame$HO ,k=neighbor)
  
  confusionMatrix <- table(predictions, test.frame$HO)
  #print(confusionMatrix)
  
  errorRate_list[i] <- (confusionMatrix[2] + confusionMatrix[3])/(sum(confusionMatrix))
  # confusion matrix is a list of 2-dim vectors, this picks out the off diagonal (error) terms
  
}

errorRate_knn <- sum(errorRate_list)/k
    
```

`r error_v_neighbor_plot`

Fitting all 3 ml prediction models, we arrive at a 10-fold cross validated error rates:

## Error Comparison Table

```{r error_comparison, echo=FALSE, warning=FALSE}

comparison_table <-  matrix(
    data = c(errorRate_logistic, "", errorRate_lda, "", errorRate_knn, paste("k =", optimalSize, sep = " ") ) , 
    nrow=2, ncol=3)
colnames(comparison_table) <- c("Logistic Regression", "LDA", "k Nearest Neighbors")
rownames(comparison_table) <- c("Average Error Rates", "Optimal Parameters")
```

`r kable(comparison_table)`

2. 
  (a) 
Simulate data where the errors are not Normal, e.g. come from a t-distribution with 3 degrees of freedom.(Assume the model is $y = 3 + 3x + \text{error}$, where there are 100 x’s spread between 0 and 3.)
  
```{r t_distribution, echo=FALSE, warning=FALSE}
#set.seed(1)

error = rt(n = 100, df = 3)
x = 1:100
for (i in 1:100){
  
  x[i] = i*3/100;
  
}
y = 3 + 3*x + error

xyplot <- plot(x,y)
```

Attached below is a plot of the simulated data from the model $y = 3 + 3x + \text{error}$ where the vector of error terms are instances drawn from a $t(3)$ distribution:

`r xyplot`

(b) Fit a linear model, and compare the standard errors of the fitted coefficients to the standard errors estimated by the bootstrap method.

```{r t_dist_error_fit, echo=FALSE, warning=FALSE}

model <- lm(y~x)
modelfit <- summary(model)
t.dist_se <- modelfit$coefficients[2,2]

modelcoefficients <- kable(modelfit$coefficients)

xy <- data.frame(x,y)

t.func = function(data,b,formula){  
# b is the random indexes for the bootstrap sample
	d = data[b,] 
# returns the beta coefficient
	return(lm(d[,1]~d[,2], data = d)$coef[2])  
	}

bootfit <- boot(data = xy, statistic = t.func, R = 100)

```

We obtain the regression model
`r modelcoefficients`
(with violated normality assumptions on the error) achieves a standard error of `r t.dist_se`, and 

```{r print_4, echo=FALSE, warning=FALSE}
(bootfit)
```

<!-- `r print(bootfit)` -->

shows that bootstraping achieves a standard error of .0148688.


(c) Now simulate where the errors are correlated. Compare the standard errors reported by the regression output (for the coefficients) to the bootstrap-estimated standard errors.

```{r correlated_errors, echo=FALSE, include=FALSE, warning=FALSE}

delta = rnorm(100,0,1)
ep = 1:100
ep[1] = delta[1]
rho = .79

for(i in 2:100){
  ep[i] = rho*ep[i-1]+delta[i] 
}

X = 1:100
for (i in 1:100){
  X[i] = i*3/100;
}
Y = 3 + 3*X + ep

XYplot <- plot(X,Y)


model.correl <- lm(Y~X)
fit.correl <- summary(model.correl)

model.correl_se <- fit.correl$coefficients[2,2]

model.correl_coefficients <- kable(fit.correl$coefficients)


XY <- data.frame(X,Y)

boot.correl <- boot(data = XY, statistic = t.func, R = 100)

```

We can simulate data $y = 3 + 3x + \epsilon$ similar to as discussed before, where the errors $\epsilon$ are correlated linearly: $\epsilon_1 = \delta_1$ follows a normal distribution about $0$, and similarly $\delta_i \sim N(0,1)$, but $\epsilon_{i+1} = \rho \epsilon_i + \delta_i$ (we pick $\rho = .79$ in the model below).

```{r echo=FALSE} 
XYplot <- plot(X,Y)

```

We obtain the regression model
`r model.correl_coefficients`
(with violated normality assumptions on the error) achieves a standard error of `r model.correl_se`, and 

```{r print_2, echo=FALSE, warning=FALSE}
(boot.correl)
```

<!-- `r print(bootfit)` -->

shows that bootstraping achieves a standard error of 0.01323802.
